Tidal streams in triaxial systems

### 2015-02-22 ms.tex
minor changes from draft reading

\title{Tidal streams in triaxial systems}
\author{Adrian M. Price-Whelan\altaffilmark{\colum,\adrn},
	    Kathryn V. Johnston\altaffilmark{\colum},
	    Monica Valluri\altaffilmark{\mich}}

@@

On the largest scales ($\gtrsim$30~kpc for a Milky-Way-like galaxy) in the halos of galaxies, gravitational potentials are thought to be triaxial. Despite suggestive evidence, this fundamental prediction from $\Lambda$CDM cosmology has not been conclusively confirmed from observations. Around other galaxies, it is generally hard to measure the 3D mass profile because they are seen in projection. From the Earth's position within the Milky Way, our view of our own halo and proximity gives us a unique chance to directly measure the 6D positions of stars and model the shape of the mass distribution at large radii. Unfortunately, the Milky Way halo has a low density of visible tracers, but luckily many of the halo stars are likely associated with debris stripped from in-falling satellite galaxies.

As a satellite galaxy or globular cluster orbits within some larger system, mass is eroded due to the tidal forces of the host galaxy potential. If the orbit of such an object is regular, only mildly eccentric, and stays far from sharp density features (e.g., the disk), the disruption is a steady process and the mass is slowly stretched into long streams of matter. The phase-space density and therefore the observable properties of these tidal streams are sensitive to the internal progenitor properties, orbit of the progenitor, and underlying mass distribution of the host galaxy. The morphological evolution of the debris depends on the spread of orbital properties (e.g., actions or frequencies) of the debris and the orbit of the progenitor system, both of which are set by the shape and radial profile of the gravitational potential of the host galaxy. In the halo of the Milky Way, tidal streams of stars are found over a large range of distances ($\sim$10-100~kpc). By simultaneously modeling the observed phase-space density of stream stars along with the potential, it is hoped that we may infer the dark matter distribution of the Milky Way on these scales.

Tidal streams are generally modeled as 1D structures formed on regular orbits with simple internal progenitor dynamics and density evolution. Methods span a range of complexity from orbit-fitting, to \emph{Streakline} or particle-spray models, to action-space density modeling, to direct $N$-body simulations. All methods have been tested in some way on simulated observations of data and authors have mainly demonstrated their ability to recover parameters of analytic, static potential forms. Many of the methods have been [...] in axisymmetric or otherwise simple potentials but claim to extend to more complex, generalized potentials (e.g., triaxial or multi-component potentials).

\citet{pearson15} tried to reproduce observations of the stellar stream density from the globular cluster Palomar 5 in a single oblate and single triaxial potential using \emph{Streakline} models \citep{kuepper12}. They use the observed SDSS number density of stars and a limited number of radial velocities measured for stream members to fit model streams to the data. In the oblate potential (a three-component, bulge, disk, and spherical halo potential), a thin model stream is easily found that reproduces the observed stellar density morphology of the stream. In the triaxial potential (the potential from \cite{law10}, a three component, bulge, disk, and triaxial halo fit to Sagitarrius stream data), the model streams generically form large, two-dimensional ``fans'' of debris near the ends, and there are no physically reasonable progenitor orbits that reproduce the observed thinness of the stream given the observational constraints of the present-day position and velocity of the cluster. This result suggests that the shape of the host galaxy potential directly influences the morphology of the resulting debris. In order to extend tidal stream models to more generic potentials, it is important to understand why this stream-fanning occurs and whether this and other strange debris morphologies are generic to more complex potentials.

One obvious difference between the two potentials considered in \citet{pearson15} is the extra symmetry of the oblate potential. For an integrable Hamiltonian, $H$, with $N$ degrees of freedom (dof), the equations of motion are most simply expressed in angle-action coordinates. In these coordinates, the actions, $J_n$, are integrals of the motion, and the conjugate angle variables, $\theta_n$, increase linearly with time
\begin{align}
	\act_n &= {\rm const.}\\
	\dot{\act}_n &= -\frac{\partial H}{\partial \theta_n} = 0\\
	\theta_n(t) &= \Omega_n t + \theta_n(0)\\
	\dot{\theta}_n &= \frac{\partial H}{\partial \act_n} = \Omega_n(\act_1,\act_2,...,\act_n)\label{eq:aafreq}
\end{align}
where $i=1,2,...N$, and $\Omega_n$ are the fundamental frequencies of the orbit. Owing to the involutive property of the actions and the periodic nature of the angle variables, the topology of angle-action space is equivalent to that of the surface of an $N$-dimensional torus, and thus orbits are often referred to in terms of \emph{orbital tori}: each set of actions, $(\act_1,\act_2,...\act_N)$, uniquely determine the torus associated with a given orbit and the angles specify the location on the surface of the $N$-torus. If a global transformation to angle-action coordinates exists, the potential is said to be ``integrable.''

If to first order galactic halos are triaxial, galactic gravitational potentials will almost certainly not be globally integrable. If a triaxial potential is close to integrable such that the Hamiltonian can be written in terms of a small perturbation, $\epsilon$, away from some integrable Hamiltonian, $H_0$,
\begin{equation}
	H(\bs{\theta}, \bs{\act}) = H_0(\bs{\act}) + \epsilon H_1(\bs{\theta}, \bs{\act})
\end{equation}
then a large number of regular orbits will survive the perturbation (by way of the KAM theorem)\footnote{3-vectors are represented with bold symbols, e.g., $\bs{J}=(J_1,J_2,J_3)$.}. The tori that survive will in general be separated by regions of irregular or chaotic motion, and thus any transformations to angle-action coordinates can only be defined locally. Analytic transformations to angle-action coordinates are only known for a handful of simple, integrable potentials. Much effort has gone in to developing general methods for transforming either from position-velocity to angle-action coordinates, $(\bs{x},\bs{v})\rightarrow(\bs{\theta},\bs{J})$ \citep{many}, or vice-versa, $(\bs{\theta},\bs{J})\rightarrow(\bs{x},\bs{v})$ \citep{many}.

@@

Resonant orbits are rare \citep{merritt99} but dynamically important in systems with several degrees of freedom as they determine the structure of orbit-space. Stable, resonant orbits generate families of orbits that may behave or appear similarly, whereas unstable resonances are typically associated with regions of chaos. However not all orbits can be trivially associated with resonances due to the infinitely complex, hierarchical structure of resonances (though the importance or ``size'' of a resonance falls off inversely with the norm of its integer vector, $|\bs{m}|^{-\alpha}$, with $\alpha > 0$). In autonomous, two dof systems, stochastic regions are generally surrounded and separated by regular motion, thus chaotic orbits are confined. In three or more dof the chaotic regions will connect and overlap. This is the origin of Arnold diffusion \citep{arnold64}, which allows chaotic orbits to radically change shape by traversing the web of unstable resonances and are only strictly bounded to their own energy hypersurface. This phenomenon is perhaps easier to visualize in action-space: in two dof systems, if a given set of tori are destroyed due to potential perturbations, an orbit in this now chaotic region is still bounded between the surrounding tori due to energy conservation. In three dimensions, the extra degree of freedom allows chaotic orbits to ``escape'' this confinement.

For small potential perturbations, many regular orbits survive and only small chaotic islands are introduced. As the strength of the perturbation increases, eventually all tori associated with conditionally-periodic motion will be destroyed. Next, the uni-resonant tori are destroyed, and finally the bi-resonant tori -- these are least susceptible to destruction from perturbations \citep[e.g., see][]{valluri98}. However, even orbits that are strictly chaotic may behave regularly over timescales much longer than, for example, the dynamical time of the system. Chaotic can therefore be loosely classified by being either \emph{strongly} or \emph{weakly} chaotic: strongly chaotic orbits behave irregularly over a few ($\sim$10) dynamical times, whereas weakly chaotic orbits behave regularly over this timescale (we quantify this distinction with the experiments of Section~\ref{sec:experiments}). This rough classification is tied to local measures, such as the Lyapunov time (Section~\ref{sec:lyap}) or frequency diffusion time (Section~\ref{sec:naff}).

@@

The density distributions within dark matter halos formed in cosmological N-body simulations are typically triaxial \citep[e.g.,][]{jing02, bett07, zemp09, veraciro11}. \citet{jing02} found that a triaxial generalization of the classic NFW density profile \citep{navarro96} generates excellent fits to their high-resolution N-body simulations, and they provide probability distributions for the axis ratios of the many halos formed in their simulations. They find median axis ratios of $c/a \approx 0.55$ and $b/a \approx 0.77$ where $a$ is the major axis, $b$ the intermediate, and $c$ the minor axis.\footnote{Note that \citet{jing02} use the opposite notation so that $c$ is the major and $a$ is the minor axis.} They find significant scatter in the distributions of concentration parameter, $c_e$, or scale radius (depending on choice of parametrization).

@@

For this work, we use the triaxial potential expression from \citet{leesuto03}, parametrized in a slightly different manner. In terms of spherical coordinates\footnote{$(r,\phi,\theta)$ = (radius, azimuth, colatitude)} with the radius normalized by the scale radius, $u = r/r_s$
\begin{align}
	\Phi(u,\phi,\theta) &\approx \frac{v_c^2}{A}\left[F_1(u) + \frac{1}{2}(e_b^2 + e_c^2)F_2(u) + \frac{1}{2} [(e_b\sin\theta \sin\phi)^2 + (e_c\cos\theta)^2] F_3(u) \right]\\
	A &= \left(\ln2 - \frac{1}{2}\right) + \left(\ln2-\frac{3}{4}\right) (e_b^2 + e_c^2)
\end{align}
where $e_b = \sqrt{1 - (b/a)^2}$, $e_c = \sqrt{1 - (c/a)^2}$, and $v_c$ is the circular velocity at the scale radius, $r_s$, for the spherical case. The functions $F_i(u)$ are given in the appendix of \cite{leesuto03}. We chose $r_s=40~{\rm kpc}$ and $v_c = 150~{\rm km}~{\rm s}^{-1}$ by taking the mean halo concentration found by \cite{jing02} for a ${\rm M}_{vir} \approx 10^{12}~\msun$ halo, $c_e\approx5$, and assuming $R_{vir}\approx200~{\rm kpc}$. Figure~\ref{fig:potential} shows the rotation curve of this potential along the three principal axes.

@@

This potential is a simple and unrealistic model for the total potential of a Milky-Way-like galaxy, however it represents a conservative choice for exploring the effect of chaos in the halos of such galaxies. Real galactic potentials will have a significant component due to the disk and bulge and may have twisting inertia tensors (cite), radially changing axis ratios \citep[e.g.,][]{veraciro11}, significant substructure \citep{zemp09}, or time dependence (either from bulk rotation, mass growth, mergers, etc.; cite). We expect all of these effects to increase the amount and influence of chaos; see the Discussion (Section~\ref{sec:discussion}) for a few simple demonstrations.

@@

We use the Dormand-Prince 8th-order Runge-Kutta scheme (TODO: cite Dormand 1981) to integrate orbits in the above potential. In particular, we use the Python wrapper to the FORTRAN routines provided in the Scipy package \citep{scipy}. For all orbits, we ensure that energy is conserved to $\Delta E/E_0 \leq 10^{-9}$ by the end of integration. Unless otherwise specified, orbits are integrated for $\sim$200 orbital periods with 500 steps per orbital period. Some box orbits require more internal integration steps in order to conserve energy to the required level.

@@

The simplest and perhaps most well-known method for assessing chaotic motion is to analyze the Lyapunov spectrum or maximum Lyapunov exponent of an orbit. Defining the vector $\bs{w} = (q_1,...,q_N,p_1,...,p_N)$ for some set of canonical coordinates $(q_i,p_i)$, we can write Hamilton's equations as 
\begin{equation}
	\dot{\bs{w}} = \bs{\mathcal{J}} \frac{\partial H}{\partial \bs{w}} = f(\bs{w}) \label{eq:ham}
\end{equation}
where $\bs{\mathcal{J}}$ is the $2N \times 2N$ canonical Poisson tensor (also called the symplectic matrix) defined by
\begin{equation}
	\bs{\mathcal{J}} = \left( \begin{array}{c:c} 0 & \ident \\ \hdashline -\ident & 0 \end{array} \right)
\end{equation}
with $N$-dimensional identity matrices $\ident$. If we consider a nearby position separated from $\bs{w}$ by an infinitesimal deviation $\delta \bs{w}$, such that $\bs{w}' = \bs{w} + \delta \bs{w}$. We can expand to linear order in the deviation about the parent orbit and write the equations of motion for the deviation as follows
\begin{align}
	\dot{\bs{w}}' &= f(\bs{w}')\\
	\dot{\bs{w}} + \dot{\delta \bs{w}} &= f(\bs{w} +  \delta \bs{w})\\
	\dot{\bs{w}} + \dot{\delta \bs{w}} &\approx f(\bs{w}) + \jac_{\bs{w}}\cdot\delta \bs{w} + \mathcal{O}(\delta \bs{w}^2)\\
	\dot{\delta \bs{w}} &\approx \jac_{\bs{w}} \cdot  \delta \bs{w}\label{eq:deviate}
\end{align}
where $\jac$ is the Jacobian of Eq.~\ref{eq:ham}, evaluated at the parent orbit. For chaotic orbits, the maximum eigenvalue of the solution matrix to Eq.~\ref{eq:deviate} is positive real, leading to exponential divergence of nearby orbits.

Computing the Lyapunov spectrum for a given orbit is often not necessary if one is only interested in characterizing the degree of chaos. Instead, it is often more efficient to compute the maximum Lyapunov exponent by estimating the finite-time maximum Lyapunov exponent (FTMLE), defined as 
\begin{equation}
	l_{\rm max}(t) = \frac{1}{t}\ln \frac{\|\delta \bs{w}(t)\|}{\|\delta \bs{w}_0\|}.\label{eq:lmax}
\end{equation}
The maximum Lyapunov exponent is the limit as $t\rightarrow \infty$ of the FTMLE
\begin{equation}
	\lambda_{\rm max} = \lim_{t\rightarrow\infty}l_{\rm max}(t). \label{eq:lyapmax}
\end{equation}
Numerically computing this quantity is not trivial because (1) obviously the limit to infinity is not possible and (2) the norm of the deviation vector $\|\delta \bs{w}(t)\|$ is expected to increase exponentially for chaotic orbits, leading to numerical problems. To circumvent these issues, it is sufficient to instead start a nearby orbit with some small initial deviation with norm $\delta_0$, integrate for a sufficiently small amount of time, $\tau$, then renormalize the deviation back to the initial norm (e.g., cite Bennetin et al. 1976, Tabor 1989). 
%The following pseudocode outlines this procedure:\footnote{see Gary for Python implementation?}\\
%\begin{algorithmic}[1]
%\State {\bf define} orbital integration timestep, $h$, and number of steps, $K$
%\State {\bf define} initial norm of deviation vector, $\delta_0$, to be sufficiently small
%\State {\bf define} renormalization integration period, $\tau \ll hK$ 
%\For{each timestep when integrating the main orbit, $\bs{w}(t)$}
%\State step forward the orbit and deviation vector orbit by one timestep, $t_{i-1} \rightarrow t_i$
%\If {a normalization timestep}
%\State measure and store the length of the deviation vector, $\delta_i = \|\delta \bs{w}_i\|$
%\State renormalize the length of the deviation vector, $\delta \bs{w}_i = \delta \bs{w}_i (\delta_0/\delta_i)$
%\EndIf
%\EndFor 
%\end{algorithmic}
The FTMLE after a given number of timesteps, $N$, is then estimated as
\begin{equation}
	l_N = \frac{1}{N\tau}\sum_i^N \ln \, \delta_i
\end{equation}
and the MLE is
\begin{equation}
	\lambda_{\rm max} = \lim_{N\rightarrow \infty} l_N.
\end{equation}

For most regular orbits, deviations will grow close to linearly or as some power-law of time. 
%As we have seen (Section~\ref{sec:introduction}), if the orbit is regular, there exists a local transformation to action-angle variables where the angle variables increase linearly with time, $\theta_n \propto \Omega_n t$. We can look at small variations around the orbit defined by the above equations:\footnote{Note the implied summation in repeated indices in the above expression.}
%\begin{align}
%	\frac{d (\delta \theta_n)}{dt} &= \frac{\partial \Omega_n}{\partial J_m} \delta J_m
%\end{align}
%These equations are trivially integrated:
%\begin{align}
%	\delta \theta_n(t) &= \delta \theta_n(0) + \left(\frac{\partial \Omega_n}{\partial J_m} \delta J_m \right) t
%\end{align}
%It's clear then that the norm of the deviation vector grows linearly with time:
%\begin{align}
%	\|\delta \bs{w}(t)\| &= \left[\sum_n (\delta \theta_n(t))^2 + \sum_n (\delta J_n)^2\right]^{1/2}\\
%	&= \left[\sum_n \left(\delta \theta_n(0) + \frac{\partial \Omega_n}{\partial J_m} \delta J_m t\right)^2 + \sum_n (\delta J_n)^2\right]^{1/2}\\
%	&\propto t.
%\end{align}
From Eq.~\ref{eq:lmax} and Eq.~\ref{eq:lyapmax} it is evident that any deviation vector that grows as a power law with time, $t^k$, will asymptote to 0 from the limit 
\begin{equation}
	\lambda_{\rm max} \propto \lim_{t\rightarrow \infty} k \frac{\ln t}{t} = 0.
\end{equation}
At long times, the implication is that the the numerically computed MLE should approach 0 like $t^{-1}$ for regular orbits. For chaotic orbits, the divergence is exponential, and the limit should converge to the rate of the exponential: the Lyapunov exponent.

@@

[Figure: for same orbits as in Lyapunov section, show frequency diffusion by computing frequencies in rolling windows?]

@@

\subsection{Adding a disk?}

>>>>

% Highlight that stochastic tubes have not been studied much...

\title{Tidal streams in triaxial systems}
\author{Adrian M. Price-Whelan\altaffilmark{\colum,\adrn},
	    Kathryn V. Johnston\altaffilmark{\colum},
	    Monica Valluri\altaffilmark{\mich}}
\date{\centering \today}

@@

On the largest scales ($\gtrsim$30~kpc for a Milky-Way-like galaxy) in the halos of galaxies, gravitational potentials are thought to be triaxial. Despite suggestive evidence from a range of complimentary techniques, this fundamental prediction from $\Lambda$CDM cosmology has not been conclusively verified by observations. Around other galaxies, it is generally hard to measure the 3D mass profile because they are seen in projection. From the Earth's position within the Milky Way, our view of our own halo and proximity gives us a unique chance to directly measure the 6D positions of stars and model the shape of the mass distribution at large radii. Unfortunately, the Milky Way halo has a low density of visible tracers, but luckily many of the halo stars are likely associated with debris stripped from in-falling satellite galaxies.

As a satellite galaxy or globular cluster orbits within some larger system, mass is eroded due to the tidal forces of the host galaxy potential. If the orbit of such an object is regular, only mildly eccentric, and stays far from sharp density features (e.g., the disk), the disruption is a steady process and the mass is slowly stretched into long streams of matter. The phase-space density and therefore the observable properties of these tidal streams are sensitive to the internal progenitor properties, orbit of the progenitor, and underlying mass distribution of the host galaxy. The morphological evolution of the debris depends on the spread of orbital properties (e.g., actions or frequencies) of the debris and the orbit of the progenitor system, both of which are also set by the shape and radial profile of the gravitational potential of the host galaxy. In the halo of the Milky Way, tidal streams of stars are found over a large range of distances ($\sim$10-100~kpc). By simultaneously modeling the observed phase-space density of stream stars along with the potential, it is hoped that we may infer the dark matter distribution of the Milky Way on these scales.

Tidal streams are generally modeled as 1D structures formed on regular orbits with simple internal progenitor dynamics and density evolution. Methods span a range of complexity from orbit-fitting, to \emph{Streakline} or particle-spray models, to action-space density modeling, to direct $N$-body simulations. All methods have been tested in some way on simulated observations of data and authors have mainly demonstrated their ability to recover parameters of analytic, static potential forms. Many of the methods have been shown to work in axisymmetric or otherwise simple potentials but claim to extend to more complex, generalized potentials (e.g., triaxial or multi-component potentials).

\citet{pearson15} tried to reproduce observations of the stellar stream density from the globular cluster Palomar 5 in a single oblate and single triaxial potential using \emph{Streakline} models \citep{kuepper12}. They use the observed SDSS number density of stars and a limited number of radial velocities measured for stream members to fit model streams to the data. In the oblate potential (a three-component, bulge, disk, and spherical halo potential), a thin model stream is easily found that reproduces the observed stellar density morphology of the stream. In the triaxial potential (the potential from \cite{law10}, a three component, bulge, disk, and triaxial halo fit to Sagitarrius stream data), the model streams generically form large, two-dimensional ``fans'' of debris near the ends, and there are no physically reasonable progenitor orbits that reproduce the observed thinness of the stream given the observational constraints of the present-day position and velocity of the cluster. This result suggests that the shape of the host galaxy potential directly influences the morphology of the resulting debris. In order to extend tidal stream models to more generic potentials, it is important to understand why this stream-fanning occurs and whether this and other strange debris morphologies are generic to more complex potentials. This also has implications for modeling the accretion histories of galaxies. Debris morphology is often separated into two classes --- streams (usually from high angular momentum progenitor orbits) and shells (from radially plunging orbits) --- but in practice the distinction may not be as clear due to other mixing effects.

The obvious difference between the two potentials considered by \citet{pearson15} is the extra symmetry of the oblate potential. For an integrable Hamiltonian, $H$, with $N$ degrees of freedom (dof), the equations of motion are most simply expressed in angle-action coordinates \cite[e.g.,][]{goldstein80}. In these coordinates, the actions, $J_n$, are integrals of the motion, and the conjugate angle variables, $\theta_n$, increase linearly with time
\begin{align}
	\dot{\act}_n &= -\frac{\partial H}{\partial \theta_n} = 0\\
	\act_n &= {\rm const.}\\
	\dot{\theta}_n &= \frac{\partial H}{\partial \act_n} = \Omega_n(\act_1,\act_2,...,\act_n)\label{eq:aafreq}\\
	\theta_n(t) &= \Omega_n t + \theta_n(0)
\end{align}
where $i=1,2,...N$, and $\Omega_n$ are the fundamental frequencies of the orbit. The topology of angle-action space is equivalent to that of the surface of an $N$-dimensional torus, and thus orbits are often referred to in terms of \emph{orbital tori}: each set of actions, $(\act_1,\act_2,...\act_N)$, uniquely ``label'' the torus associated with a given orbit and the angles specify the location on the surface of the $N$-torus. If a global transformation to angle-action coordinates exists, the potential is said to be \emph{integrable}.

If galactic halos are triaxial, galactic gravitational potentials will almost certainly not be globally integrable. If a triaxial (three dof) potential is close to integrable such that the Hamiltonian can be written in terms of a small perturbation, $\epsilon$, away from some integrable Hamiltonian, $H_0$,
\begin{equation}
	H(\bs{\theta}, \bs{\act}) = H_0(\bs{\act}) + \epsilon H_1(\bs{\theta}, \bs{\act})
\end{equation}
then a large number of regular orbits will survive the perturbation (by way of the KAM theorem)\footnote{3-vectors are represented with bold symbols, e.g., $\bs{J}=(J_1,J_2,J_3)$.}. The tori that survive will in general be separated by regions of irregular or chaotic motion, and thus any transformations to angle-action coordinates can only be defined locally. Analytic transformations to angle-action coordinates are only known for a handful of simple, integrable potentials. Much effort has gone in to developing general methods for transforming either from position-velocity to angle-action coordinates, $(\bs{x},\bs{v})\rightarrow(\bs{\theta},\bs{J})$ \citep{many}, or vice-versa, $(\bs{\theta},\bs{J})\rightarrow(\bs{x},\bs{v})$ \citep{many}.

@@

Resonant orbits are rare \citep{merritt99} but dynamically important in systems with several degrees of freedom as they determine the structure of orbit-space. Stable, resonant orbits generate families of orbits that may behave or appear similarly, whereas unstable resonances are typically associated with regions of chaos. Not all orbits can be trivially associated with resonances due to the infinitely complex, hierarchical structure of resonances (though the importance or ``size'' of a resonance falls off inversely with the norm of its integer vector, $|\bs{m}|^{-\alpha}$, with $\alpha > 0$). In autonomous, two dof systems, stochastic regions are generally surrounded and separated by regular motion, thus chaotic orbits are confined. In three or more dof the chaotic regions will connect and overlap. This is the origin of Arnold diffusion \citep{arnold64}, which allows chaotic orbits to radically change shape by traversing the web of unstable resonances and are only strictly bounded to their own energy hypersurface. This phenomenon is perhaps easier to visualize in action-space: in two dof systems, if a given set of tori are destroyed due to potential perturbations, an orbit in this now chaotic region is still bounded between the surrounding tori due to energy conservation. In three dimensions, the extra degree of freedom allows chaotic orbits to ``escape'' this confinement.

For small potential perturbations, many regular orbits survive and only small chaotic islands are introduced. As the strength of the perturbation increases, eventually all tori associated with conditionally-periodic motion will be destroyed. Next, the uni-resonant tori are destroyed, and finally the bi-resonant tori -- these are least susceptible to destruction from perturbations \citep[e.g., see][]{valluri98}. However, even orbits that are strictly chaotic may behave regularly over timescales much longer than, for example, the dynamical time of the system. Chaotic orbits can therefore be loosely classified by being either \emph{strongly} or \emph{weakly} chaotic: strongly chaotic orbits behave irregularly over a few ($\sim$10) dynamical times, whereas weakly chaotic orbits behave regularly over this timescale (we quantify this distinction with the experiments of Section~\ref{sec:experiments}). This rough classification is tied to local measures, such as the Lyapunov time (Section~\ref{sec:lyap}) or frequency diffusion time (Section~\ref{sec:naff}).

@@

The density distributions within dark matter halos formed in cosmological N-body simulations are typically triaxial \citep[e.g.,][]{jing02, bett07, zemp09, veraciro11}. \citet{jing02} found that a triaxial generalization of the classic NFW density profile \citep{navarro96} generates excellent fits to their high-resolution N-body simulations, and they provide probability distributions for the axis ratios of the many halos formed in their simulations. They find median axis ratios of $c/a \approx 0.55$ and $b/a \approx 0.77$ where $a$ is the major axis, $b$ the intermediate, and $c$ the minor axis.\footnote{Note that \citet{jing02} use the opposite notation so that $c$ is the major and $a$ is the minor axis.} These are consistent with constraints from weak lensing that place a lower limit on minor-to-major axis ratios of $c/a\gtrsim0.5$ \citep{vanuitert12}. They find significant scatter in the distributions of concentration parameter, $c_e$, or scale radius (depending on choice of parametrization).

@@

In this work, we use the triaxial potential expression from \citet{leesuto03}, parametrized in a slightly different manner. In terms of spherical coordinates\footnote{$(r,\phi,\theta)$ = (radius, azimuth, colatitude)} with the radius normalized by the scale radius, $u = r/r_s$
\begin{align}
	\Phi(u,\phi,\theta) &\approx \frac{v_c^2}{A}\left[F_1(u) + \frac{1}{2}(e_b^2 + e_c^2)F_2(u) + \frac{1}{2} [(e_b\sin\theta \sin\phi)^2 + (e_c\cos\theta)^2] F_3(u) \right]\\
	A &= \left(\ln2 - \frac{1}{2}\right) + \left(\ln2-\frac{3}{4}\right) (e_b^2 + e_c^2)
\end{align}
where $e_b = \sqrt{1 - (b/a)^2}$, $e_c = \sqrt{1 - (c/a)^2}$, and $v_c$ is the circular velocity at the scale radius, $r_s$, for the spherical case. The functions $F_i(u)$ are given in the appendix of \cite{leesuto03}. We chose $r_s=40~{\rm kpc}$ and $v_c = 150~{\rm km}~{\rm s}^{-1}$ by taking the mean halo concentration for a ${\rm M}_{vir} \approx 10^{12}~\msun$ halo, $c_e\approx5$, from \cite{jing02} and by assuming $R_{vir}\approx200~{\rm kpc}$. Figure~\ref{fig:potential} shows the rotation curve of this potential along the three principal axes.

@@

This potential is a simple and unrealistic model for the total potential of a Milky-Way-like galaxy, however it represents a conservative choice for exploring the effect of chaos in the halos of such galaxies. Realistic Falactic potentials will have a significant component due to the disk and bulge and may have twisting inertia tensors (cite), radially changing axis ratios \citep[e.g.,][]{veraciro11}, significant substructure \citep{zemp09}, or time dependence (either from bulk rotation, mass growth, mergers, etc.; cite). We expect all of these effects to increase the amount and influence of chaos; see the Discussion (Section~\ref{sec:discussion}) for a few simple demonstrations.

@@

We use the Dormand-Prince 8th-order Runge-Kutta scheme (TODO: cite Dormand 1981) to integrate orbits in the above potential. In particular, we use the Python wrapper to the FORTRAN routines provided in the Scipy package \citep{scipy}. For all orbits, we ensure that energy is conserved to $\Delta E/E_0 \leq 10^{-9}$ by the end of integration. Unless otherwise specified, orbits are integrated for $\sim$250 orbital periods with 500 steps per orbital period. Some box orbits require more internal integration steps in order to conserve energy to the required level.

@@

The simplest and perhaps most well-known method for assessing chaotic motion is to analyze the Lyapunov spectrum or maximum Lyapunov exponent (MLE) of an orbit. Defining the vector $\bs{w} = (q_1,...,q_N,p_1,...,p_N)$ for some set of canonical coordinates $(q_i,p_i)$, we can write Hamilton's equations as 
\begin{equation}
	\dot{\bs{w}} = \bs{\mathcal{J}} \frac{\partial H}{\partial \bs{w}} = f(\bs{w}) \label{eq:ham}
\end{equation}
where $\bs{\mathcal{J}}$ is the $2N \times 2N$ canonical Poisson tensor (also called the symplectic matrix) defined by
\begin{equation}
	\bs{\mathcal{J}} = \left( \begin{array}{c:c} 0 & \ident \\ \hdashline -\ident & 0 \end{array} \right)
\end{equation}
with $N$-dimensional identity matrices $\ident$. If we consider a nearby phase-space position, $\bs{w}'$, separated from $\bs{w}$ by an infinitesimal deviation, $\delta \bs{w}$, such that $\bs{w}' = \bs{w} + \delta \bs{w}$. We can expand to linear order in the deviation about the parent orbit and write the equations of motion for the deviation as follows
\begin{align}
	\dot{\bs{w}}' &= f(\bs{w}')\\
	\dot{\bs{w}} + \dot{\delta \bs{w}} &= f(\bs{w} +  \delta \bs{w})\\
	\dot{\bs{w}} + \dot{\delta \bs{w}} &\approx f(\bs{w}) + \jac_{\bs{w}}\cdot\delta \bs{w} + \mathcal{O}(\delta \bs{w}^2)\\
	\dot{\delta \bs{w}} &\approx \jac_{\bs{w}} \cdot  \delta \bs{w}\label{eq:deviate}
\end{align}
where $\jac$ is the Jacobian of Eq.~\ref{eq:ham}, evaluated at the parent orbit (the Hessian of the Hamiltonian). For chaotic orbits, the maximum eigenvalue of the solution matrix to Eq.~\ref{eq:deviate} is positive real, leading to exponential divergence of nearby orbits.

Computing the Lyapunov spectrum for a given orbit is often not necessary if one is only interested in characterizing the degree of chaos. Instead, it is often more efficient to compute the maximum Lyapunov exponent by estimating the finite-time maximum Lyapunov exponent (FTMLE), defined as 
\begin{equation}
	l_{\rm max}(t) = \frac{1}{t}\ln \frac{\|\delta \bs{w}(t)\|}{\|\delta \bs{w}_0\|}.\label{eq:lmax}
\end{equation}
The maximum Lyapunov exponent is the limit as $t\rightarrow \infty$ of the FTMLE
\begin{equation}
	\lambda_{\rm max} = \lim_{t\rightarrow\infty}l_{\rm max}(t). \label{eq:lyapmax}
\end{equation}
Numerically computing this quantity is not trivial because (1) obviously the limit to infinity is not possible and (2) the norm of the deviation vector $\|\delta \bs{w}(t)\|$ is expected to increase exponentially for chaotic orbits, leading to numerical problems. To circumvent these issues, it is sufficient to instead start a nearby orbit with some small initial deviation with norm $\delta_0$, integrate for a sufficiently small amount of time, $\tau$, then renormalize the deviation back to the initial norm (e.g., cite Bennetin et al. 1976, Tabor 1989). 
%The following pseudocode outlines this procedure:\footnote{see Gary for Python implementation?}\\
%\begin{algorithmic}[1]
%\State {\bf define} orbital integration timestep, $h$, and number of steps, $K$
%\State {\bf define} initial norm of deviation vector, $\delta_0$, to be sufficiently small
%\State {\bf define} renormalization integration period, $\tau \ll hK$ 
%\For{each timestep when integrating the main orbit, $\bs{w}(t)$}
%\State step forward the orbit and deviation vector orbit by one timestep, $t_{i-1} \rightarrow t_i$
%\If {a normalization timestep}
%\State measure and store the length of the deviation vector, $\delta_i = \|\delta \bs{w}_i\|$
%\State renormalize the length of the deviation vector, $\delta \bs{w}_i = \delta \bs{w}_i (\delta_0/\delta_i)$
%\EndIf
%\EndFor 
%\end{algorithmic}
The FTMLE after a given number of timesteps, $N$, is then estimated as
\begin{equation}
	l_N = \frac{1}{N\tau}\sum_i^N \ln \, \delta_i
\end{equation}
where $\delta_i=\|\delta \bs{w}(t_i)\|$ and the MLE is
\begin{equation}
	\lambda_{\rm max} = \lim_{N\rightarrow \infty} l_N.
\end{equation}

For most regular orbits, deviations will grow close to linearly or as some power-law of time. 
%As we have seen (Section~\ref{sec:introduction}), if the orbit is regular, there exists a local transformation to action-angle variables where the angle variables increase linearly with time, $\theta_n \propto \Omega_n t$. We can look at small variations around the orbit defined by the above equations:\footnote{Note the implied summation in repeated indices in the above expression.}
%\begin{align}
%	\frac{d (\delta \theta_n)}{dt} &= \frac{\partial \Omega_n}{\partial J_m} \delta J_m
%\end{align}
%These equations are trivially integrated:
%\begin{align}
%	\delta \theta_n(t) &= \delta \theta_n(0) + \left(\frac{\partial \Omega_n}{\partial J_m} \delta J_m \right) t
%\end{align}
%It's clear then that the norm of the deviation vector grows linearly with time:
%\begin{align}
%	\|\delta \bs{w}(t)\| &= \left[\sum_n (\delta \theta_n(t))^2 + \sum_n (\delta J_n)^2\right]^{1/2}\\
%	&= \left[\sum_n \left(\delta \theta_n(0) + \frac{\partial \Omega_n}{\partial J_m} \delta J_m t\right)^2 + \sum_n (\delta J_n)^2\right]^{1/2}\\
%	&\propto t.
%\end{align}
From Eq.~\ref{eq:lmax} and Eq.~\ref{eq:lyapmax} it is evident that any deviation vector that grows as a power law with time, $t^k$, will asymptote to 0 from the limit 
\begin{equation}
	\lambda_{\rm max} \propto \lim_{t\rightarrow \infty} k \frac{\ln t}{t} = 0.
\end{equation}
At long times, the numerically computed MLE should approach 0 as $t^{-1}$ for regular orbits. For chaotic orbits, the divergence is exponential, and the limit should converge to the rate of the exponential: the Lyapunov exponent.

@@

We have found that for ``sticky'' or mildly chaotic orbits, reliably computing the Lyapunov exponent can require integration times equivalent to thousands of orbital periods. In comparison, owing to the precision of NAFF, detecting small frequency diffusion rates can be done with only hundreds of orbital periods. [example? comparison?] 
[Figure: for same orbits as in Lyapunov section, show frequency diffusion by computing frequencies in rolling windows?]

@@
